
\section{Dynamically Composable Traits in \name}
\label{sec:traits}

\bruno{Should we say something about structural typing, purely
  functional? if so, where? here? the introduction?} 
%\name is a purely functional OO language with a polymorphic 
%structural type system based on disjoint intersection types~\cite{}.
%The choice of structural typing is simply due to simplicity, but we
%similar ideas should also work in a nominal 

This section introduces and showcases the features of \name. 
In this section we focus on \name's native support
for dynamically composable traits. We demonstrate that \name overcomes all those
limitations mentioned in Section~\ref{sec:critique}.

We describe a representation of graphical objects such as circles, ovals, or
buttons as our running example. The example is illustrative of typical
uses of JavaScript style mixins, and it is adapted from an online
tutorial~\footnote{\url{https://javascriptweblog.wordpress.com/2011/05/31/a-fresh-look-at-javascript-mixins/}}
on JavaScript mixins.
We will use traits to structure the representation and factor out reusable
components. Along the way, we also explain briefly how various features in \name
are encoded behind the scene. All code snippets in this section are runnable in
our prototype implementation.

\subsection{Simple Traits}

\name natively supports a simple, yet expressive form of
traits~\cite{scharli2003traits}.  Traits provide a mechanism of code
reuse in object-oriented programming, that can be used to model
disciplined forms of multiple inheritance. One interesting aspect
about traits is the way conflicting features that are typical arise in
multiple inheritance are dealt with. Instead of automatically resolved
by scoping rules, conflicts are detected by the type system, and
programmers are required to explicitly resolve them.

There are three interesting points about \name's traits: 1) they are
\emph{statically typed}; 2) they support \emph{constructors}; and 3)
they support \emph{dynamic inheritance}. The support for such
combination of features is one of the key novelties of \name. This, in
combination with the separation of inheritance and subtyping makes the
trait system particularly flexible and expressive.

In what follows, we demonstrate various trait features in \name. A
comparison with the traditional trait model can be found in
Section~\ref{sec:discuss}. The desugaring process from traits to
\bname is discussed in Section~\ref{sec:desugar}.

\paragraph{Specifying Traits}
A trait is a collection of related methods that characterize only
a specific perspective of the features of an object. Therefore, compared with
programs using class-based inheritance, programs using traits usually have a large number of
small traits rather than fewer but larger classes. Code reuse with traits is
easier than with classes, since traits are usually shorter and can be
\textit{composed}. It is the ease of composition that makes traits such a
appealing language feature: two traits can be freely ``added''
together. Trait composition is
a symmetric operation, and trait systems provide conflict detection.

Here is a simple trait \lstinline{point} with its two coordinates \lstinline{x}
and \lstinline{y}.
\lstinputlisting[linerange=4-8]{../examples/box.txt}% APPLY:linerange=POINT_DEF

%The syntax is very similar to Scala's, with the \lstinline{def} keyword
%beginning a term declaration.

The first thing worth noting is that there are two declarations here.
%At this point the reader may wonder what is the \lstinline{type} declaration
%(\lstinline$Point$) for. 
In mainstream OO languages such as Java, a class
declaration such as \lstinline[language=java]$class C { ... }$ does two things
at the same time:

\begin{itemize}
\item Declaring a \textit{template} for creating objects;
\item Declaring a new \textit{type}.
\end{itemize}

In contrast, trait declarations in \name only do the former. Type
declarations are defined separately. For example, the type declaration 
for \lstinline{Point} declares the types of the fields \lstinline{x}
and \lstinline{y}. Separating the two roles of
classes has advantages in terms of flexibility, although one could
argue that it is longwinded sometimes. It is easy 
enough to add some syntactic sugar to do the two roles with a single 
definition, but we will stick to separate definitions in this paper 
to illustrate its value.

The trait \lstinline{point} provides a simple example of trait
declarations in \name. The following discussions illustrate 
some fundamental features of \name's traits.

\paragraph{Traits as templates for creating objects} A first an
obvious difference of traits in \name to many other models of traits
is that they directly serve as templates for objects. In many other
trait models~\cite{}, traits are complemented by classes, which take 
the responsability for object creation. In particular, most models
of traits do not allow constructors for traits. As the
\lstinline{point} definition illustrates, however, traits in \name 
have a single constructor, which can have an arbitrary number of arguments.

\paragraph{Trait requirements versus provided functionality}
When modelling a trait there are two important 
aspects: what are the \emph{requirements} of a trait; and what is 
the functionality that a trait \emph{provides}? The requirements of a
trait denote the types/methods that the self type should support for
defining the functionality of the trait.
The trait \lstinline{point}
requires nothing, and provides implementations for the fields 
\lstinline{x} and \lstinline{y}.  In \name, the type
of \lstinline$self$ denotes what methods/types are required 
by a trait. \name uses a syntax (inspired by Scala 
self type declarations) where the self reference is explicitly named 
and can optionaly be given a type. When there is no type annotation 
on self, this means that the trait requires nothing. In this case 
the type of \lstinline{self} is the \emph{top} type. 
This is different from typical OO languages, where the default type 
of a class is the same as the class being defined.

%The purpose of
%declaring types is to use them for type annotations of the self-reference, and
%creating instances. In the trait literature, a trait usually requires a set of
%methods that server as parameters for the provided behaviour. In \name, the type
%of \lstinline$self$ denotes what methods are required.

\paragraph{Creating objects}
Creating an object is done using the \lstinline{new} keyword. A
difference to other OO languages is that the \lstinline{new} specifies
both the intended type of the object, as well how to construct the
object using constructors.
\lstinputlisting[linerange=12-12]{../examples/box.txt}% APPLY:linerange=POINT_TEST

\paragraph{Inheriting traits}
A trait can be extended by inheriting all members of
other traits, and defining additional members. This is, in some sense, similar
to class inheritance in traditional OO languages. Traits can inherit
from one or more traits, provided that
no conflicts arize. The trait \lstinline$circle$ extends
\lstinline{point} with an extra field \lstinline{radius}.
\lstinputlisting[linerange=18-21]{../examples/box.txt}% APPLY:linerange=CIRCLE_DEF

The keyword \lstinline{inherits} is one of the two options in \name to introduce
inheritance. In the above example, trait \lstinline{circle} inherits from trait
\lstinline{point} two methods \lstinline{x} and \lstinline{y} plus its own
method \lstinline{radius}.\bruno{Should we talk about the second
  option here?}

\paragraph{Intersection types model subtyping}
The type \lstinline{Circle} is defined as the intersection of type
\lstinline{Point} and a record type with the field
\lstinline{radious}. An intersection
type~\cite{dunfield2014elaborating}~\bruno{Dunfield did not invent
  intersection types. Put the appropriate reference here.} such as \lstinline{A & B} contains exactly
those values which can be used as values of type \lstinline{A} and of type
\lstinline{B}. As such, \lstinline{A & B} immediately introduces a subtyping
relation between itself and its two component types \lstinline{A} and
\lstinline{B}. In the previous example, \lstinline{Circle} is, unsurprisingly, a subtype of
\lstinline{Point}.

\paragraph{Inheritance and subtyping are separated}
Now it is clear that, unlike the common covariant
model, \name separates the concept of subtyping from inheritance. Those two
concepts are not necessarily entangled, although in this particular example,
inheritance goes along with subtyping.

\subsection{Traits with Requirements and ``Abstract'' Methods}

So far the traits we have seen have no requirements. However, very often it is
necessary to call methods using the self reference. When this happens
we need to express explicit requirements on the traits. An interesting aspect
of \name's trait model is that there is no need for \emph{abstract
  methods}. Instead, abstract methods can be simulated as requirements 
of a trait. 

In our example, each graphical object can be decomposed into two aspects -- its
geometry and its functions. In case of circles, we already have a trait
\lstinline{circle} representing its geometry. We now proceed to define its
functions by another trait.
\lstinputlisting[linerange=29-38]{../examples/box.txt}% APPLY:linerange=CIRCLE_FNS

\noindent Note how in \lstinline$circleFns$ the type of the self-reference is
\lstinline$Circle$, because the definition of the functions in the
trait requires the \lstinline{radius} definition. Note also that
the trait itself does not actually contains a \emph{radius}
definition. In many other OO models a similar program could be achieved
by having an \emph{abstract} radius definition. In \name there are no
abstract definitions (method or fields), but we can express a
requirement on the trait to achieve a similar result. 

Requirements 
are satisfied at object creation time.
When \lstinline$circleFns$ is instantiated and
composed with other traits, it must be composed with an implementation of
\lstinline$Circle$. For example:
\lstinputlisting[linerange=42-42]{../examples/box.txt}% APPLY:linerange=CIRCLE_FULL

The object \lstinline{circleWithFns} enables us to call
methods from different traits on a single object, such as
\lstinline{circleWithFns.area()} and \lstinline{circleWithFns.radius}.

\paragraph{Composition of traits}
The definition \lstinline{circleWithFns} also shows the second option to introduce inheritance, namely by
\textit{composition} of traits. Composition of traits is denoted by
the operator \lstinline{&}.  
Thus \name offers two options when using inheritance: we can either compose beforehand
when declaring traits (using \lstinline{inherits}), or compose at the object
creation point (using \lstinline{new} and \lstinline{&} operator).

%Under the hood, inheritance is accomplished by using the \textit{merge operator}
%(denoted by \lstinline{,,}). The merge operator~\cite{dunfield2014elaborating}
%allows two arbitrary values to be merged, with the resulting type being an
%intersection type. 
%For example the type of \lstinline{2 ,, true} is
%\lstinline{Int & Bool}.

\paragraph{Multiple trait inheritance} To further demonstrate multiple
trait inheritance, consider adding buttons. The trait of buttons and its type is:
\lstinputlisting[linerange=47-52]{../examples/box.txt}% APPLY:linerange=BUTTON_DEF

Similarly we define functions for buttons as well.
\lstinputlisting[linerange=56-63]{../examples/box.txt}% APPLY:linerange=BUTTON_FNS

\noindent To make a round button from the existing traits the trait \lstinline{roundButton} just inherits everything from
traits \lstinline{circle} and \lstinline{button}:
\lstinputlisting[linerange=67-69]{../examples/box.txt}% APPLY:linerange=ROUNDBUTTON_DEF

\noindent Finally we create a round button on the fly and test its functionality
as follows:
\lstinputlisting[linerange=125-128]{../examples/box.txt}% APPLY:linerange=ROUNDBUTTON_TEST

\subsection{Detecting And Resolving Conflicts in Trait Composition}
\label{sec:conflicts}

A common problem in multiple inheritance are conflicts. For example,
when inheriting from two traits that have the same field, then it is
unclear which implementation to chose. There are various approaches to
deal with conflicts. The trait-based approach requires conflicts to be
resolved at the level of the composition by the programmers, otherwise
the program is rejected by the type system. 

The following example
shows how conflicting methods in two traits are detected and resolved.
Let us introduce an oval shape.
\lstinputlisting[linerange=74-77]{../examples/box.txt}% APPLY:linerange=ASOVAL_DEF

The following trait gets rejected because both \lstinline{asOval} and
\lstinline{circle} have a conflicting \lstinline{radius} method.
\lstinputlisting[linerange=82-83]{../examples/box.txt}% APPLY:linerange=CONFLICT_DEF

\noindent As mentioned in Section~\ref{} \name's type system is based
on intersection types. More concretely, \name uses a type system based
on \emph{disjoint intersection types}~\cite{oliveira2016disjoint}. Disjointness, in its simplest
form, means that the set of values of both types are disjoint. The above
problematic program fails at passing type checking precisely because both of the
traits (\lstinline{circle} and \lstinline{asOval}) have the same type for the
\lstinline{radius} method, thus violating the disjointness conditions.

\paragraph{Resolving conflicts}
To resolve the conflict, the programmer needs to explicitly state which
\lstinline{radius} gets to stay. \name provides such a means , the so-called
\textit{exclusion} operator (denoted by \lstinline{\}), which allows one to
avoid a conflict before it occurs. The following is one choice, and is accepted
by \name again.
\lstinputlisting[linerange=89-90]{../examples/box.txt}% APPLY:linerange=CONFLICT_RESOLVE
\bruno{As I mentioned before, this exclusion operator is not very
  adequate for a system with disjoint intersection types, since our
  notion of disjointness is not purely based on names: we account for
  types as well! So restriction should take into account the types of
  the fields.}

\subsection{Dynamic Instantiation}

One difference with traditional traits or classes is that traits in \name are
quite dynamic: we are able to compose traits \textit{dynamically} and then
instantiate them later. This is impossible in traditional OO language, such as
Java, since classes being instantiated must be known statically. In \name, as we
will explain in Section~\ref{sec:desugar}, traits are just terms: they are
first-class values and can be passed around or returned from a function.

Let us extend the functions of circles by another method \lstinline{inCircle},
which, given a point, tests if the point lies inside the circle.
\lstinputlisting[linerange=105-108]{../examples/box.txt}% APPLY:linerange=CIRCLE_FNS2

\lstinline{CircleFns2} just extends \lstinline{CircleFns} by another method
\lstinline{inCircle}, which, as shown in trait \lstinline{circleFns2}, is
implemented by invoking the \lstinline{norm} method from something of type
\lstinline{Circle & Norm}. So what is this \lstinline{Norm} type? It consists of
a single method \lstinline{norm} -- distance of a point to the origin . We
provide two different norms via two traits.
\lstinputlisting[linerange=94-101]{../examples/box.txt}% APPLY:linerange=NORM_DEF

To facilitate creating round buttons with different norms baked in, we define a
factory that takes a trait \lstinline$norm$ and produces a round button:
\lstinputlisting[linerange=112-114]{../examples/box.txt}% APPLY:linerange=POINT_FUNC

Here \lstinline{norm}, which is a trait used in the object creation of
\lstinline{roundButtonFac}, is \emph{parametrized}. To express the
type of such unknown trait \name has a special 
\lstinline{Trait} type constructor, and \lstinline$Trait[Point, Norm]$
denotes the type of traits that conforms to the \lstinline$Norm$ type with dependency
on \lstinline{Point}. This gives us the
flexibility to choose different norms at \emph{runtime}. For example, below is another version of
round button that provides \lstinline{inCircle} method with Euclidean norm baked
in.
\lstinputlisting[linerange=118-118]{../examples/box.txt}% APPLY:linerange=ROUNDBUTTON_TEST2

\bruno{This tour doesn't go over polymorphism!}